---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
After unzipping the file, we read the data in and assign the correct class to each column. The dates are conveniently stored in the default form YYYY-MM-DD so we don't need any special arguments to convert that column to class Date. 
```{r}
unzip("activity.zip")
raw_data <- read.csv("activity.csv", header=TRUE, stringsAsFactors=FALSE,
                          colClasses=c("integer", "Date", "integer"))
```
Let's see what the data looks like:
```{r}
head(raw_data)
```

## What is the mean total number of steps taken per day?
We're told that we can ignore the missing values for this part.
It appears that the only column to have missing values is the steps column:
```{r}
any(is.na(raw_data$steps))
any(is.na(raw_data$date))
any(is.na(raw_data$interval))
```

For this question, since we're told to ignore missing values, we'll keep only the complete cases:
```{r}
completedata <- raw_data[complete.cases(raw_data), ]
```
To calculate the total number of steps taken per day we want to take the
sum of the steps column, grouped by date.
```{r}
totalsteps <- tapply(completedata$steps, completedata$date, sum)
```


Here's a histogram of the total number of steps taken each day. I 
played with the "breaks" argument until I liked the histogram.
```{r}
hist(totalsteps, breaks=seq(0, 25000, by=2500),
     xlab="Total Steps", main="Total Number of Steps Taken Each Day")
```

Finally, here are the mean and mean and median of the total number of steps taken per day:
```{r}
mean(totalsteps)
median(totalsteps)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
